# Practical: Linear Regression

In this practical you will go through some of the basics of linear modelling in `R` as well as simulating data. The practical contains the following elements:

- simulate linear regression model
- investigate parameters
- characterise prediction accurary
- correlation of real world data

We will use `reshape2`, `ggplo2`, and `bbmle` pcakges. Run the following command to make sure they are installed and loaded

```{r eval=FALSE}
install.packages("ggplot2")
install.packages("reshape2")
install.packages("bbmle")
```

```{r}
library(ggplot2)
library(reshape2)
library(bbmle)
```

## Simulating data

You will simulate data based on the simple linear regression model:

$$
y_i = \beta_0 + \beta_1\, x_i + \epsilon_i,
$$

where $(x_i, y_i)$ represent the $i$-th measurement pair with $i = 1, \ldots, N$, $\beta_0$ and $\beta_1$ are regression coefficients representing intercept and slope respectively. We assume the noise term $\epsilon_i \sim N(0, \sigma^2)$ is normally distributed with zero mean and variance $\sigma^2$.

First we define the values of the parameters of linear regression $(\beta_0, \beta_1, \sigma^2)$:

```{r }
b0 <- 10 # regression coefficient for intercept
b1 <- -8 # regression coefficient for slope
sigma2 <- 0.5 # noise variance
```

In the next step we will simulate $N = 100$ covariates $x_i$ by randomly sampling from the standard normal distribution:

```{r }
set.seed(198) # set a seed to ensure data is reproducible
N <- 100 # no of data points to simulate
x <- rnorm(N, mean = 0, sd = 1) # simulate covariate

```

Next we simulate the error term:

```{r }
# simulate the noise terms, rnorm requires the standard deviation
e <- rnorm(N, mean = 0, sd = sqrt(sigma2))
```

Finally we have all the parameters and variables to simulate the response variable $y$:

```{r }
# compute (simulate) the response variable
y = b0 + b1 * x + e
```

We will plot our data using `ggplot2` so the data need to be in a `data.frame` object:

```{r sim-data}
# Set up the data point
sim_data <- data.frame(x = x, y = y)

# create a new scatter plot using ggplot2
ggplot(sim_data, aes(x = x, y = y)) +
  geom_point()

```

We define the true data `y_true` to be the true linear relationship between the covariate and the response without the noise.

```{r }
# Compute true y values
y_true <- b0 + b1 * x

# Add the data to the existing data frame
sim_data$y_true <- y_true

```

Now we will add the true values of $y$ to the scatter plot:

```{r scatter-true}
lr_plot <- ggplot(sim_data, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(x = x, y = y_true), colour = "red")

print(lr_plot)
```

## Fitting simple linear regression model

### Least squared estimation

Now that you have simulated data you can use it to regress $y$ on $x$, since this is simulated data we know the parameters and can make a comparison. In `R` we can use the function `lm()` for this, by default it implements a least squares estimate:

```{r }
# Use the lm function to fit the data
ls_fit <- lm(y ~ x, data = sim_data)

# Display a summary of fit
summary(ls_fit)
```

The output for `lm()` is an object (in this case `ls_fit`) which contains multiple variables. To access them there are some built in functions, e.g. `coef()`, `residuals()`, and `fitted()`. We will explore these in turn:

```{r lm-fit}
# Extract coefficients as a named vector
ls_coef <- coef(ls_fit)

print(ls_coef)

# Extract intercept and slope
b0_hat <- ls_coef[1] # alternative ls_fit$coefficients[1]
b1_hat <- ls_coef[2] # alternative ls_fit$coefficients[2]

# Generate the predicted data based on estimated parameters
y_hat <- b0_hat + b1_hat * x
sim_data$y_hat <- y_hat # add to the existing data frame

# Create scatter plot and lines for the original and fitted
lr_plot <- ggplot(sim_data, aes(x = x, y = y)) +
  geom_point() +
  geom_line(aes(x = x, y = y_true), colour = "red", size = 1.3) +
  # plot predicted relationship in blue
  geom_line(aes(x = x, y = y_hat), colour = "blue")

# force Rstudio to display the plot
  print(lr_plot)
```

The estimated parameters and the plot shows a good correspondence between fitted regression parameters and the true relationship between $y$ and $x$. We can check this by plotting the residuals, this data is stored as the `residuals` parameter in the `ls_fit` object.

```{r resid-scatter}
# Residuals
ls_residual <- residuals(ls_fit) # can also be accessed via ls_fit$residuals

# scatter plot of residuals
plot(ls_residual)
```

A better way of summarising the data is to visualise them as a histogram:

```{r resid_hist}

hist(ls_residual)

```

We expect the mean and variance of the residuals to be close to the level used to generate the data.

```{r }
print(mean(ls_residual))

print(var(ls_residual))
```

This is as expected since subtracting a good fit from the data leaves $\epsilon$ which has $0$ mean and $0.5$ variance.

### Maximum likelihood estimation

Next you will look at maximum likelihood estimation based on the same data you simulated earlier. This is a bit more involved as it requires you to explicitly write the function you wish to minimise. The function we use is part of the `bbmle` package.

```{r warning=FALSE}
# Loading the required package
library(bbmle)

# function that will be minimised. It takes as arguments all parameters
# Here we are helped by the way R works we don't have to explicitly pass x.
# The function will use the existing estimates in the environment
mle_ll <- function(beta0, beta1, sigma) {
  # first we predict the response variable based on the guess for our response
  y_pred = beta0 + beta1 * x

  # next we calculate the normal distribution based on the predicted value
  # the guess for sigma and return the log
  log_lh <- dnorm(y, mean = y_pred, sd = sigma, log = TRUE)

  #  We returnr the negative sum of the log likelihood
  return(-sum(log_lh))
}

# This is the function that actually performs the estimation
# The first variable here is the function we will use
# The second variable passed is a list of initial guesses of parameters
mle_fit <- mle2(mle_ll, start = list(beta0 = -1, beta1 = 20, sigma = 10))

# With the same summary function as above we can output a summary of the fit
summary(mle_fit)
```

The estimated parameters using the maximum likelihood are also a very good estimate of the true values.

## Effect of variance

Now investigate the quality of the predictions further by simulating more data sets and seeing how the variance affects the quality of the fit as indicated by the mean-squared error (mse).

To start you will define some parameter for the simulations, the number of simulations to run for each variance, and the variance values to try.

```{r }
# number of simulations for each noise level
n_simulations <- 100

# A vector of noise levels to try
sigma_v <- c(0.1, 0.4, 1.0, 2.0, 4.0, 6.0, 8.0)
n_sigma <- length(sigma_v)

# Create a matrix to store results
mse_matrix <- matrix(0, nrow = n_simulations, ncol = n_sigma)

# name row and column
rownames(mse_matrix) <- c(1:n_simulations)
colnames(mse_matrix) <- sigma_v
```

Next we will write a nested `for` loop. The first loop will be over the variances and a second loop over the number of repeats. We will simulate the data, perform a fit with `lm()`. We can use the `fitted()` function on the resulting object to extract the fitted values $\hat{y}$ and use this to compute the mean-squared error from the true value $y$.

```{r }
# loop over variance
for (i in 1:n_sigma) {
  sigma2 <- sigma_v[i]

  # for each simulation
  for (it in 1:n_simulations) {

    # Simulate the data
    x <- rnorm(N, mean = 0, sd = 1)
    e <- rnorm(N, mean = 0, sd = sqrt(sigma2))
    y <- b0 + b1 * x + e

    # set up a data frame and run lm()
    sim_data <- data.frame(x = x, y = y)
    lm_fit <- lm(y ~ x, data = sim_data)

    # compute the mean squared error between the fit and the actual y's
    y_hat <- fitted(lm_fit)
    mse_matrix[it, i] <- mean((y_hat - y)^2)

  }
}
```

We created a matrix to store the mse values, but to plot them using `ggplot2` we have to convert them to a `data.frame`. This can be done using the `melt()` function form the `reshape2` library. We can compare the results using boxplots.


```{r effect-plot, fig.width = 12}
library(reshape2)

# convert the matrix into a data frame for ggplot2
mse_df <- melt(mse_matrix)
# rename the columns
names(mse_df) <- c("Simulation", "variance", "MSE")

# now use a boxplot to look at the relationship between
# mean-squared prediction error and variance
mse_plt <- ggplot(mse_df, aes(x = variance, y = MSE)) +
  geom_boxplot(aes(group = variance))

print(mse_plt)
```

You can see that the variances of the mse and the value of the mse go up with increasing variance in the simulation.

What changes do you need to make to the above function to plot the accuracy of the estimated regression coefficients as a function of variance?

## Exercise

### Part I

Read in the data in `storks.txt`, compute the correlation and comment on it.

The data represents `no of storks` (column 1) in Oldenburg Germany from $1930 - 1939$ and the number of people (column 2).

### Part II

Fit a simple linear model to the two data sets supplied (`lr_data1.Rdata` and `lr_data2.Rdata`). In both files the $(x,y)$ data is saved in two vectors, $x$ and $y$.

Download the data from Canvas, you can read it into `R` and plot it with the following commands:

```{r ex-1}
load("lr_data1.Rdata")
plot(x, y)

load("lr_data2.Rdata")
plot(x, y)

```

Fit the linear model and comment on the differences between the data.


### Part III

Investigate how the sample size will affect the quality of the fit using mse, use the code for investigating the affect of variance as inspiration.



